Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-07-28 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Minor changes

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (ACC and DLPFC)';

@@

### Identify gene outliers
Set outlier expression of genes in each samples to NA
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA

# Distribution of fraction NA genes in samples
tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
writeLines('Distribution of fraction NAs in samples')
quantile(tmp.frac, seq(0, 1, 0.1)) %>%
  round(3) 
top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
```
Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Function to convert NA values to zero
changeNAtoZero <- function(mtrx){
  mtrx[is.na(mtrx)] = 0
  return(mtrx)
}

# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
CQN.GENE_EXPRESSION$E[is.na(PROCESSED_COUNTS$filteredExprMatrix$counts)] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, scale.=T, center = T)

@@

# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)

@@

tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION %>% changeNAtoZero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

```{r temp1, include=F}
dev.off()
gc()
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
                         levels = colnames(VOOM.SVA.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.SVA.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
names(DE)[7:8] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r DM1_OBJ$properties$id`       |
|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                 parentId = "syn9872607",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)
library(sva)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn9872607";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

@@

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

@@

# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)

@@

tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",

                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",

                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",

                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### Source Code
[R markdown](`r thisFile`)
